---
title: "Supplementary Material"
author: "[John Zobolas](https://github.com/bblodfon)"
date: last-modified
description: "Supporting information for paper 'Examining  properness in the external validation of survival models with squared and logarithmic losses'"
format:
  html:
    date: last-modified
    code-block-bg: true
    code-copy: true
    code-fold: true
    code-overflow: wrap
    code-block-border-left: true
    toc: true
    toc-location: left
    html-math-method: katex
    page-layout: full
execute:
  freeze: true
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

## High variability for generated survival and censoring distributions {-}

To illustrate the diversity of the Weibull distributions used in our simulation design, we plot both the density $f(t)$ and survival $S(t)$ functions for 10 randomly sampled shape $\alpha$ and scale $\sigma$ parameter pairs.
These distributions represent either the true survival, predicted survival, or censoring times across simulations.
The wide parameter range induces substantial variability, ensuring a rich family of distributions and increasing the sensitivity of our evaluation to detect violations of properness.
By construction, all data-generating mechanisms assume $Y$ independent of $C$, satisfying the random censoring assumption.

```{r, message=FALSE}
#| fig-cap: "Weibull density (top) and survival (bottom) curves for randomly generated shape, α, and scale, σ parameters. Each curve corresponds to a different parameter pair, illustrating the broad variability across simulated distributions."

library(ggplot2)
library(patchwork)
library(dplyr)

set.seed(20250408)

shape = runif(10, 0.5, 5)
scale = runif(10, 0.5, 5)

dw = mapply(function(shape,scale) dweibull(seq.int(1,5,0.1), shape,scale),shape,scale) |>
  reshape2::melt() |>
  mutate(x = rep(seq.int(1,5,0.1), 10),
        group = rep(sprintf("(%.2f,%.2f)",shape,scale), each = 41)) |>
  select(x, group, value)
pw = mapply(function(shape,scale) pweibull(seq.int(1,5,0.1), shape,scale, lower.tail=F),shape,scale) |>
  reshape2::melt() |>
  mutate(x = rep(seq.int(1,5,0.1), 10),
        group = rep(sprintf("(%.2f,%.2f)",shape,scale), each = 41)) |>
  select(x, group, value)

(ggplot(dw, aes(x = x, y = value, group = group, color = group)) + ylab("f(x)")) /
(ggplot(pw, aes(x = x, y = value, group = group, color = group)) + ylab("S(x)")) +
  plot_layout(guides='collect') &
  geom_line() &
  theme_classic() &
  guides(color = guide_legend(title='(α,σ)')) &
  scale_color_viridis_d(option = "turbo")

# ggsave("weibull_plots.png", device = png, dpi=600, width =7, height = 4)
```

## Table 1 {-}

:::{.callout-note}
This section supports the "Simulation Experiments" section of the paper.

Use [properness_test.R](https://github.com/survival-org/scoring-rules-2024/blob/main/properness_test.R) to run the experiments.
To run for different sample sizes `n`, use [run_tests.sh](https://github.com/survival-org/scoring-rules-2024/blob/main/run_tests.sh).
Merged simulation results for all $n$ are available [here](https://github.com/survival-org/scoring-rules-2024/blob/main/results/res_sims10000_distrs1000_0.rds).
:::

Load libraries and explore data:
```{r, message=FALSE}
library(dplyr)
library(tibble)
library(tidyr)
library(DT)

res = readRDS("results/res_sims10000_distrs1000_0.rds")

res[sample(1:nrow(res), 10), ] |>
  DT::datatable(rownames = FALSE, options = list(searching = FALSE)) |>
  formatRound(columns = 3:40, digits = c(rep(3,8), rep(5,30)))
```

Each row corresponds to one simulation, using $m = 1000$ draws from Weibull survival distributions at a fixed sample size $n \in {10, \dots, 10\,000}$.

The output columns are:

- `sim`: simulation index ($k$ in the paper)
- `n`: sample size
- `n_events`: number of uncensored events observed in the simulated dataset
- `max_t`: maximum observed follow-up time
- `m_true`, `m_pred`: true and predicted mean survival times from the generating Weibull models
- `m_true_est`, `m_pred_est`: estimated mean survival times using the trapezoidal rule on the observed time point support
- `S_Y_q10`, `S_Y_med`, `S_Y_q90`: values of the true event-time survival function $S_Y(t)$ evaluated at the 10th percentile, median, and 90th percentile of the observed time points.
- `S_C_q10`, `S_C_med`, `S_C_q90`: Values of the censoring survival function $S_C(t)$ evaluated at the 10th percentile, median, and 90th percentile of the observed time points.
- `S_Y_tail`, `S_C_tail`: tail survival probability at `max_t` for event and censoring distributions
- `epsilon`: $\epsilon = S_Y(t_{max}) \times S_C(t_{max})$
- `{surv|pred|cens|}_{scale|shape}` the scale and shape of the Weibull distributions for true, predicted and censoring times respectively
- `prop_cens`: proportion of censoring in each simulation
- `tv_dist`: the total variation distance between the true and the predicted Weibull distribution (closer to 0 means more similar distributions)
- `{score}_diff`, `{score}_sd`: the mean and standard deviation of the score difference $D$ between true and predicted distributions across draws

Scoring rules analyzed:

- The **SBS** at the 10th, 50th and 90th percentiles of observed times
- The **ISBS**, using 50 equidistant points between the 5th and 80th percentile of observed times
- The **RCLL** and its weighted version **wRCLL** (or **RCLL\***)

We compute 95% confidence intervals (CIs) for the mean score difference $D$ (true - predicted) using a $t$-distribution.
A violation is marked statistically significant if:

- The mean score difference exceeds a positive threshold: $\bar{D} > 0.0001$
- The CI excludes zero (i.e., $\text{CI}_\text{lower} > 0$)

```{r}
res = res |>
  select(!matches("shape|scale|prop_cens|tv_dist|n_events|max_t|m_true|m_pred"))

measures = c("SBS_q10", "SBS_median", "SBS_q90", "ISBS", "RCLL", "wRCLL")
n_distrs = 1000 # `m` value from paper's experiment
tcrit = qt(0.975, df = n_distrs - 1) # 95% t-test
threshold = 1e-4

data = measures |>
  lapply(function(m) {
    mean_diff_col = paste0(m, "_diff")
    mean_sd_col = paste0(m, "_sd")
    
    res |>
      select(sim, n, !!mean_diff_col, !!mean_sd_col, starts_with("shift_"), 
             epsilon, starts_with("S_")) |>
      select(-ends_with("tail")) |>
      mutate(
        se = .data[[mean_sd_col]] / sqrt(n_distrs),
        CI_lower = .data[[mean_diff_col]] - tcrit * se,
        CI_upper = .data[[mean_diff_col]] + tcrit * se,
        signif_violation = .data[[mean_diff_col]] > threshold & CI_lower > 0,
        # Keep relevant SBS columns
        shift_q10 = if (m == "SBS_q10") shift_q10 else NA_real_,
        S_Y_q10   = if (m == "SBS_q10") S_Y_q10 else NA_real_,
        S_C_q10   = if (m == "SBS_q10") S_C_q10 else NA_real_,
        shift_med = if (m == "SBS_median") shift_med else NA_real_,
        S_Y_med   = if (m == "SBS_median") S_Y_med else NA_real_,
        S_C_med   = if (m == "SBS_median") S_C_med else NA_real_,
        shift_q90 = if (m == "SBS_q90") shift_q90 else NA_real_,
        S_Y_q90   = if (m == "SBS_q90") S_Y_q90 else NA_real_,
        S_C_q90   = if (m == "SBS_q90") S_C_q90 else NA_real_
      ) |>
      select(!se) |>
      mutate(metric = !!m) |>
      relocate(metric, .after = 1) |>
      rename(diff = !!mean_diff_col,
             sd   = !!mean_sd_col)
  }) |>
  bind_rows()

data$metric = factor(
  data$metric,
  levels = measures,
  labels = c("SBS (Q10)", "SBS (Median)", "SBS (Q90)", "ISBS", "RCLL", "RCLL*")
)
```

For example, the results for the first simulation and $n = 250$ across all scoring rules show no violations of properness:
```{r}
data |> 
  filter(sim == 1, n == 250) |>
  select(-starts_with("shift"), -epsilon, -starts_with("S_")) |>
  DT::datatable(rownames = FALSE, options = list(searching = FALSE)) |>
  formatRound(columns = 4:7, digits = 5)
```

We summarize violations across simulations (`sim`), by computing for each score & sample size:

- Number of significant violations
- Violation rate
- Average, median, min and max score difference among simulations where violations occurred

```{r}
all_stats = 
  data |> 
  group_by(n, metric) |> 
  summarize(
    n_violations = sum(signif_violation),
    violation_rate = mean(signif_violation),
    diff_mean = if (any(signif_violation)) mean(diff[signif_violation]) else NA_real_,
    diff_median = if (any(signif_violation)) median(diff[signif_violation]) else NA_real_,
    diff_min = if (any(signif_violation)) min(diff[signif_violation]) else NA_real_,
    diff_max = if (any(signif_violation)) max(diff[signif_violation]) else NA_real_,
    .groups = "drop"
  )

all_stats |>
  arrange(metric) |>
  DT::datatable(
    rownames = FALSE,
    options = list(searching = FALSE),
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: center; font-size:150%",
      "Table 1: Empirical violations of properness")
  ) |>
  formatRound(columns = 4:8, digits = 5)
```

:::{.callout-tip}
## Summary

- **RCLL** and **RCLL\*** show practically no violations across any simulation or sample size.
- **SBS** shows time- and sample-size-dependent violations, mostly at smaller $n$, especially for early evaluation times ($\tau = q_{0.1}$).
All differences are small (typically < 0.01).
- **ISBS** showed minor violations only at $n < 50$; none for $n > 100$.
:::

## Quantifying Impropriety in the Survival Brier Score (SBS)

The theoretical shift from the global minimum $x^*$ of the expected SBS (i.e. the bias) is:

$$x^* - S_Y(\tau^*) = - \frac{\epsilon (1-S_Y(\tau^*))}{S_C(\tau^*) - \epsilon}$$

Filter SBS simulation data for plotting:
```{r}
plot_data = data |>
  filter(grepl("^SBS", metric)) |>
  pivot_longer(
    cols = starts_with("shift_"),
    names_to = "shift_type",
    values_to = "shift",
    values_drop_na = TRUE  # only keep the relevant shift per metric
  ) |>
  mutate(
    S_Y = case_when(
      metric == "SBS (Q10)" ~ S_Y_q10,
      metric == "SBS (Median)" ~ S_Y_med,
      metric == "SBS (Q90)" ~ S_Y_q90
    ),
    S_C = case_when(
      metric == "SBS (Q10)" ~ S_C_q10,
      metric == "SBS (Median)" ~ S_C_med,
      metric == "SBS (Q90)" ~ S_C_q90
    )
  ) |>
  select(-S_Y_q10, -S_Y_med, -S_Y_q90, -S_C_q10, -S_C_med, -S_C_q90)
```

```{r}
#| fig-cap: "Relationship between survival tail product ε and theoretical shift from the global minimum, using a sample of data from the Weibull simulation experiments"
#| cache: true
#| fig-width: 9
#| fig-height: 5

ann_df = data.frame(
  metric = unique(plot_data$metric)[1],
  x = 0.01,
  y = -0.07,
  xend = 0.01,
  yend = -0.175,
  label = "more improper"
)

# shift vs epsilon
set.seed(42)
plot_data |> 
  slice_sample(prop = 0.2) |> # 3 SBS metrics x 10 sample sizes (n) x 1000 simulations per n
  ggplot(aes(x = epsilon, y = shift, color = factor(n))) +
    geom_point(alpha = 0.5, size = 0.1) +
    facet_wrap(~metric, scales = "fixed") +
    scale_color_discrete(
      name = "Sample size (n)",
      breaks = rev(sort(unique(plot_data$n)))  # reverse legend
    ) +
    scale_x_continuous(
      limits = c(0, 0.11), 
      breaks = c(0, 0.025, 0.05, 0.075, 0.1)
    ) +
    labs(
      x = expression(epsilon == S[Y](t[max]) %*% S[C](t[max])),
      y = expression(x^"*" - S[Y](tau^"*")), # shift from optimal prediction
      color = "Sample size (n)"
    ) +
    theme_bw(base_size = 14, base_family = "Arial") +
    theme(axis.text.x = element_text(angle = 35, hjust = 1)) +
    guides(color = guide_legend(override.aes = list(size = 4))) +
    geom_segment(
      data = ann_df,
      aes(x = x, y = y, xend = xend, yend = yend),
      inherit.aes = FALSE,linewidth = 1,
      arrow = arrow(length = unit(0.25, "cm"))
    ) +
    geom_text(
      data = ann_df,
      aes(x = x + 0.01, y = y - 0.05, label = label),
      inherit.aes = FALSE,
      angle = 90,
      size = 6,
      fontface = "bold"
    )
```

```{r}
#| fig-cap: "Relationship between true survival at τ* and theoretical shift from the global minimum, using a sample of data from the Weibull simulation experiments"
#| cache: true
#| fig-width: 9
#| fig-height: 5

set.seed(42)
plot_data |> 
  slice_sample(prop = 0.2) |> # 3 SBS metrics x 10 sample sizes (n) x 1000 simulations per n
  ggplot(aes(x = S_Y, y = shift, color = factor(n))) +
  geom_point(alpha = 0.5, size = 0.3) +
  facet_wrap(~metric, scales = "fixed") +
  scale_color_discrete(name = "Sample size (n)") +
  scale_x_continuous(
    limits = c(0, 1), 
    breaks = c(0, 0.25, 0.5, 0.75, 1)
  ) +
  labs(
    x = expression(S[Y](tau^"*")),
    y = expression(x^"*" - S[Y](tau^"*"))
  ) +
  theme_bw(base_size = 14) +
  theme(panel.spacing = unit(1, "lines")) +
  guides(color = guide_legend(override.aes = list(size = 4))) +
  geom_segment(
    data = ann_df,
    aes(x = x, y = y, xend = xend, yend = yend),
    inherit.aes = FALSE,linewidth = 1,
    arrow = arrow(length = unit(0.25, "cm"))
  ) +
  geom_text(
    data = ann_df,
    aes(x = x + 0.1, y = y - 0.05, label = label),
    inherit.aes = FALSE,
    angle = 90,
    size = 6,
    fontface = "bold"
  )
```

```{r}
#| fig-cap: "Relationship between censoring survival at τ* and theoretical shift from the global minimum, using a sample of data from the Weibull simulation experiments"
#| cache: true
#| fig-width: 9
#| fig-height: 5

set.seed(42)
plot_data |> 
  slice_sample(prop = 0.2) |> # 3 SBS metrics x 10 sample sizes (n) x 1000 simulations per n
  ggplot(aes(x = S_C, y = shift, color = factor(n))) +
  geom_point(alpha = 0.5, size = 0.3) +
  facet_wrap(~metric, scales = "fixed") +
  scale_color_discrete(name = "Sample size (n)") +
  scale_x_continuous(
    limits = c(0, 1), 
    breaks = c(0, 0.25, 0.5, 0.75, 1)
  ) +
  labs(
    x = expression(S[C](tau^"*")),
    y = expression(x^"*" - S[Y](tau^"*"))
  ) +
  theme_bw(base_size = 14) +
  theme(panel.spacing = unit(1, "lines")) +
  guides(color = guide_legend(override.aes = list(size = 4))) +
  geom_segment(
    data = ann_df,
    aes(x = x, y = y, xend = xend, yend = yend),
    inherit.aes = FALSE,linewidth = 1,
    arrow = arrow(length = unit(0.25, "cm"))
  ) +
  geom_text(
    data = ann_df,
    aes(x = x + 0.1, y = y - 0.05, label = label),
    inherit.aes = FALSE,
    angle = 90,
    size = 6,
    fontface = "bold"
  )
```

## Table D1 {-}

:::{.callout-note}
The only difference with the previous experiment is that $S_C(t)$ is now being estimated using the marginal Kaplan-Meier model via `survival::survfit()` instead of using the true Weibull censoring distribution (see [helper.R](https://github.com/survival-org/scoring-rules-2024/blob/main/helper.R)).
For SBS/ISBS we use *constant interpolation* of the censoring survival distribution $S_C(t)$.
For RCLL\* we use *linear interpolation* of $S_C(t)$ to mitigate density estimation issues, i.e. for $f_C(t)$.

Use [properness_test.R](https://github.com/survival-org/scoring-rules-2024/blob/main/properness_test.R) to run the experiments.
To run for different sample sizes `n`, use [run_tests.sh](https://github.com/survival-org/scoring-rules-2024/blob/main/run_tests.sh), changing `estimate_cens` to `TRUE`.
Merged simulation results for all `n` are available [here](https://github.com/survival-org/scoring-rules-2024/blob/main/results/res_sims10000_distrs1000_1.rds).
:::

Load results (same output columns as in the previous section):
```{r}
res = readRDS("results/res_sims10000_distrs1000_1.rds") |>
  select(!matches("shape|scale|prop_cens|tv_dist|n_events|max_t|m_true|m_pred")) # remove columns
```

As before, we compute 95% confidence intervals for the score differences across $m = 1000$ draws per simulation:
```{r}
measures = c("SBS_q10", "SBS_median", "SBS_q90", "ISBS", "RCLL", "wRCLL")
n_distrs = 1000 # `m` value from paper's experiment
tcrit = qt(0.975, df = n_distrs - 1) # 95% t-test
threshold = 1e-4

data = measures |>
  lapply(function(m) {
    mean_diff_col = paste0(m, "_diff")
    mean_sd_col = paste0(m, "_sd")
    
    res |>
      select(sim, n, !!mean_diff_col, !!mean_sd_col, starts_with("shift_"), 
             epsilon, starts_with("S_")) |>
      select(-ends_with("tail")) |>
      mutate(
        se = .data[[mean_sd_col]] / sqrt(n_distrs),
        CI_lower = .data[[mean_diff_col]] - tcrit * se,
        CI_upper = .data[[mean_diff_col]] + tcrit * se,
        signif_violation = .data[[mean_diff_col]] > threshold & CI_lower > 0,
        # Keep relevant SBS columns, just in case
        shift_q10 = if (m == "SBS_q10") shift_q10 else NA_real_,
        S_Y_q10   = if (m == "SBS_q10") S_Y_q10 else NA_real_,
        S_C_q10   = if (m == "SBS_q10") S_C_q10 else NA_real_,
        shift_med = if (m == "SBS_median") shift_med else NA_real_,
        S_Y_med   = if (m == "SBS_median") S_Y_med else NA_real_,
        S_C_med   = if (m == "SBS_median") S_C_med else NA_real_,
        shift_q90 = if (m == "SBS_q90") shift_q90 else NA_real_,
        S_Y_q90   = if (m == "SBS_q90") S_Y_q90 else NA_real_,
        S_C_q90   = if (m == "SBS_q90") S_C_q90 else NA_real_
      ) |>
      select(!se) |>
      mutate(metric = !!m) |>
      relocate(metric, .after = 1) |>
      rename(diff = !!mean_diff_col,
             sd   = !!mean_sd_col)
  }) |>
  bind_rows()

data$metric = factor(
  data$metric,
  levels = measures,
  labels = c("SBS (Q10)", "SBS (Median)", "SBS (Q90)", "ISBS", "RCLL", "RCLL*")
)
```

Lastly, we summarize the significant violations across sample sizes and scoring rules:
```{r}
all_stats = 
  data |> 
  group_by(n, metric) |> 
  summarize(
    n_violations = sum(signif_violation),
    violation_rate = mean(signif_violation),
    diff_mean = if (any(signif_violation)) mean(diff[signif_violation]) else NA_real_,
    diff_median = if (any(signif_violation)) median(diff[signif_violation]) else NA_real_,
    diff_min = if (any(signif_violation)) min(diff[signif_violation]) else NA_real_,
    diff_max = if (any(signif_violation)) max(diff[signif_violation]) else NA_real_,
    .groups = "drop"
  )

all_stats |>
  arrange(metric) |>
  DT::datatable(
    rownames = FALSE,
    options = list(searching = FALSE),
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: center; font-size:150%",
      "Table D2: Empirical violations of properness using G(t)")
  ) |>
  formatRound(columns = 4:8, digits = 5)
```

:::{.callout-tip}
## Summary

Estimating $S_C(t)$ via Kaplan-Meier produced nearly identical results to using the true distribution, with no violations for RCLL/RCLL* and only small-sample effects for SBS and ISBS.
:::

## Real-World Benchmark Evaluation {-}

:::{.callout-note}
To run the experiment use [real_data_benchmark.R](https://github.com/survival-org/scoring-rules-2024/blob/main/real_data_benchmark.R).
:::

We assess the performance using three survival scoring rules (RCLL, RCLL\*, ISBS), D-calibration and Harrell's C-index on real-world survival datasets using the `mlr3proba` framework, tested with 10 repetitions of 5-fold cross-validation on the following learners:

- Kaplan-Meier (non-parametric)
- Cox proportional hazards model
- Random survival forests (RSF)

Below we show the aggregated performance metrics across all tasks and learners:
```{r}
res = readRDS("results/real_data_bm.rds")

res |>
  dplyr::mutate(
    learner_id = dplyr::recode(
      learner_id,
      "surv.kaplan" = "KM",
      "surv.ranger" = "RSF",
      "surv.coxph"  = "CoxPH"
    )
  )|>
  DT::datatable(rownames = FALSE, filter = "top", options = list(searching = TRUE)) |>
  DT::formatRound(columns = 3:7, digits = 3) |>
  DT::formatSignif(columns = 6, digits = 3) # scientific notation for surv.dcalib
```

:::{.callout-tip}
## Summary

Our experiment demonstrates substantial divergences in model assessment depending on the chosen metric (especially in the case of the scoring rules).

For instance, on datasets such as `veteran` and `whas`, which exhibit non-proportional hazards, RSF is more suitable to model non-linear effects compared to the CoxPH.
In these cases, RCLL* identifies RSF as superior, whereas RCLL and ISBS still prefer CoxPH — suggesting potential practical limitations of the latter metrics in such settings.

Additionally, RCLL* appears to better capture both discrimination and calibration.
In the `gbcs` dataset, CoxPH achieves the best RCLL, ISBS and C-index scores, yet shows extremely poor calibration ($\text{D-calib} \approx 15.7$), while RCLL* ranks Kaplan-Meier as the best — possibly reflecting better-calibrated risk estimates.
This discrepancy becomes even more pronounced in the `lung` dataset, where CoxPH again yields high RCLL/C-index/ISBS performance despite severe calibration failure ($\text{D-calib} \approx 23.5$), whereas RCLL* selects RSF instead.

These findings highlight a critical need for further research: **not all metrics labeled as "proper" in theory behave equally in practice.**
:::

## R session info {-}

```{r}
utils::sessionInfo()
```
