---
title: "Supplementary Material"
author: "[John Zobolas](https://github.com/bblodfon)"
date: last-modified
description: "Supporting information for paper: 'When Are Scoring Rules Proper? Bridging Theory and Practice in Survival Model Evaluation'"
format:
  html:
    date: last-modified
    code-block-bg: true
    code-copy: true
    code-fold: true
    code-overflow: wrap
    code-block-border-left: true
    toc: true
    toc-location: left
    html-math-method: katex
    page-layout: full
execute:
  freeze: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

Load libraries:
```{r, message=FALSE}
library(dplyr)
library(tibble)
library(ggplot2)
library(patchwork)
library(akima)
library(purrr)
library(tidyr)
library(data.table)
library(DT)
```

## Weibull distribution variability {-#weib}

To illustrate the diversity of the Weibull distributions used in our simulation design, we plot both the density $f(t)$ and survival $S(t)$ functions for 10 randomly sampled shape $\alpha$ and scale $\sigma$ parameter pairs.
These distributions represent either the true survival, predicted survival, or censoring times across simulations.
The wide parameter range induces substantial variability, ensuring a rich family of distributions and increasing the sensitivity of our evaluation to detect violations of properness.
By construction, all data-generating mechanisms assume $Y$ independent of $C$, satisfying the random censoring assumption.

```{r, message=FALSE}
#| fig-cap: "Weibull density (top) and survival (bottom) curves for randomly generated shape, $\\alpha$, and scale, $\\sigma$ parameters. Each curve corresponds to a different parameter pair, illustrating the broad variability across simulated distributions."
set.seed(0L)

shape = runif(9, 0.5, 5)
scale = runif(9, 0.5, 5)

time_grid = seq(0, 6, 0.1)

params = tibble(
  id = seq_along(shape),
  shape = shape,
  scale = scale,
  group = sprintf("(%0.2f, %0.2f)", shape, scale)
)

grid = tidyr::expand_grid(params, t = time_grid)

dw = grid |>
  mutate(value = dweibull(t, shape, scale), type = "Density f(t)")

pw = grid |>
  mutate(value = pweibull(t, shape, scale, lower.tail = FALSE), type = "Survival S(t)")

plot_df = bind_rows(dw, pw) |>
  mutate(type = factor(type, levels = c("Density f(t)", "Survival S(t)")))

ggplot(plot_df, aes(x = t, y = value, color = group, group = group)) +
  geom_line(linewidth = 0.8, alpha = 0.9) +
  facet_wrap(~ type, ncol = 1, scales = "free_y") +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(breaks = 0:6) +
  labs(
    x = expression("Time"),
    y = NULL,
    color = expression("(" * alpha * "," * sigma * ")"),
    title = "Weibull Density and Survival Curves"
  ) +
  theme_bw(base_size = 16) +
  theme(
    legend.position = "right",
    strip.text = element_text(face = "bold")
  )
#ggsave("weibull_plots.png", device = png, dpi = 600, width = 8, height = 5)
```

## Table 1 {-#tab1}

:::{.callout-note}
Use [properness_test.R](https://github.com/survival-org/scoring-rules-2024/blob/main/properness_test.R) to run the experiments.

Example execution: `Rscript properness_test.R 10000 1000 10000 FALSE 42`.

The above arguments correspond to (in that order): $K = 10\,000$ simulations, $m = 1\,000$ distributions per simulation and $n = 10\,000$ samples drawn from each distribution.
`FALSE` means that $S_C(t)$ is used directly from the true Weibull distribution and not estimated, $42$ is just a seed number we used for reproducibility.

To run for different sample sizes `n`, use [run_tests.sh](https://github.com/survival-org/scoring-rules-2024/blob/main/run_tests.sh).

Merged simulation results for all sample sizes are available [here](https://github.com/survival-org/scoring-rules-2024/blob/main/results/res_sims10000_distrs1000_0.rds).
:::

Results from 10 randomly sampled simulations:
```{r, message=FALSE}
res = readRDS("results/res_sims10000_distrs1000_0.rds")

res[sample(1:nrow(res), 10), ] |>
  DT::datatable(
    rownames = FALSE,
    options = list(dom = "t", searching = FALSE, lengthChange = FALSE)
  ) |>
  formatRound(columns = 3:40, digits = c(rep(3,8), rep(5,30)))
```

Each row corresponds to one simulation, and the quantites are averaged across $m = 1000$ draws from Weibull survival distributions at a fixed sample size $n \in \{10, \dots, 10\,000\}$.

The output columns are:

- `sim`: simulation index ($k$ in the paper)
- `n`: sample size
- `n_events`: number of uncensored events observed in the simulated dataset
- `max_t`: maximum observed follow-up time
- `m_true`, `m_pred`: true and predicted mean survival times from the generating Weibull models
- `m_true_est`, `m_pred_est`: estimated mean survival times using the trapezoidal rule on the observed time point support
- `S_Y_q10`, `S_Y_med`, `S_Y_q90`: values of the true event-time survival function $S_Y(t)$ evaluated at the 10th percentile, median, and 90th percentile of the observed time points.
- `S_C_q10`, `S_C_med`, `S_C_q90`: Values of the censoring survival function $S_C(t)$ evaluated at the 10th percentile, median, and 90th percentile of the observed time points.
- `S_Y_tail`, `S_C_tail`: tail survival probability at `max_t` for event and censoring distributions
- `epsilon`: $\hat\varepsilon = \frac{\sum_{d=1}^{1000} \left[S_Y^d(t_{\\max}^d) \times S_C^d(t_{\max}^d)\right]}{1000}$
- `{surv|pred|cens|}_{scale|shape}` the scale and shape of the Weibull distributions for true, predicted and censoring times respectively
- `prop_cens`: proportion of censoring in each simulation
- `tv_dist`: the total variation distance between the true and the predicted Weibull distribution (closer to 0 means more similar distributions)
- `{score}_diff`, `{score}_sd`: the mean and standard deviation of the score difference $D$ between true and predicted distributions across draws

Scoring rules analyzed:

- The **SBS** at the 10th, 50th and 90th percentiles of observed times
- The **ISBS**, using 50 equidistant points between the 5th and 80th percentile of observed times
- The **RCLL**
- An experimental weighted version of RCLL, which we called **wRCLL** (or **RCLL\***), see [weighted_RCLL.R](https://github.com/survival-org/scoring-rules-2024/blob/main/weighted_RCLL.R) for more details. This is not a proper scoring rule in general (at least in theory), and was included in initial versions of this work as a possible (even more proper) extension of RCLL.

:::{.callout-important}
We compute 95% confidence intervals (CIs) for the mean score difference $\bar{D}$ (true - predicted) using a $t$-distribution.
A violation is marked statistically significant if:

- The mean score difference exceeds a positive threshold: $\bar{D} > 0.0001$
- The CI excludes zero (i.e., $\text{CI}_\text{lower} > 0$)
:::

```{r}
res = res |>
  select(!matches("shape|scale|prop_cens|tv_dist|n_events|max_t|m_true|m_pred"))

measures = c("SBS_q10", "SBS_median", "SBS_q90", "ISBS", "RCLL", "wRCLL")
n_distrs = 1000 # `m` value from paper's experiment
tcrit = qt(0.975, df = n_distrs - 1) # 95% t-test
threshold = 1e-4

data = measures |>
  lapply(function(m) {
    mean_diff_col = paste0(m, "_diff")
    mean_sd_col = paste0(m, "_sd")

    res |>
      select(sim, n, !!mean_diff_col, !!mean_sd_col, starts_with("shift_"),
             epsilon, starts_with("S_")) |>
      select(-ends_with("tail")) |>
      mutate(
        se = .data[[mean_sd_col]] / sqrt(n_distrs),
        CI_lower = .data[[mean_diff_col]] - tcrit * se,
        CI_upper = .data[[mean_diff_col]] + tcrit * se,
        signif_violation = .data[[mean_diff_col]] > threshold & CI_lower > 0,
        # Keep relevant SBS columns
        shift_q10 = if (m == "SBS_q10") shift_q10 else NA_real_,
        S_Y_q10   = if (m == "SBS_q10") S_Y_q10 else NA_real_,
        S_C_q10   = if (m == "SBS_q10") S_C_q10 else NA_real_,
        shift_med = if (m == "SBS_median") shift_med else NA_real_,
        S_Y_med   = if (m == "SBS_median") S_Y_med else NA_real_,
        S_C_med   = if (m == "SBS_median") S_C_med else NA_real_,
        shift_q90 = if (m == "SBS_q90") shift_q90 else NA_real_,
        S_Y_q90   = if (m == "SBS_q90") S_Y_q90 else NA_real_,
        S_C_q90   = if (m == "SBS_q90") S_C_q90 else NA_real_
      ) |>
      select(!se) |>
      mutate(metric = !!m) |>
      relocate(metric, .after = 1) |>
      rename(diff = !!mean_diff_col,
             sd   = !!mean_sd_col)
  }) |>
  bind_rows()

data$metric = factor(
  data$metric,
  levels = measures,
  labels = c("SBS (Q10)", "SBS (Median)", "SBS (Q90)", "ISBS", "RCLL", "RCLL*")
)
```

For example, the results for the first simulation and $n = 250$ across all scoring rules show no violations of properness:
```{r}
data |>
  filter(sim == 1, n == 250) |>
  select(-starts_with("shift"), -epsilon, -starts_with("S_")) |>
  DT::datatable(
    rownames = FALSE,
    options = list(dom = "t", searching = FALSE, lengthChange = FALSE)
  ) |>
  formatRound(columns = 4:7, digits = 5)
```

We summarize violations across simulations (`sim`), by computing for each score & sample size:

- Number of significant violations
- Violation rate
- Average, median, min and max score difference among simulations where violations occurred

```{r}
all_stats =
  data |>
  group_by(n, metric) |>
  summarize(
    n_violations = sum(signif_violation),
    violation_rate = mean(signif_violation),
    diff_mean = if (any(signif_violation)) mean(diff[signif_violation]) else NA_real_,
    diff_median = if (any(signif_violation)) median(diff[signif_violation]) else NA_real_,
    diff_min = if (any(signif_violation)) min(diff[signif_violation]) else NA_real_,
    diff_max = if (any(signif_violation)) max(diff[signif_violation]) else NA_real_,
    .groups = "drop"
  )

all_stats |>
  arrange(metric) |>
  DT::datatable(
    rownames = FALSE,
    options = list(searching = FALSE),
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: center; font-size:150%",
      "Table 1: Empirical violations of properness"
    )
  ) |>
  formatRound(columns = 4:8, digits = 5)
```

:::{.callout-tip}
## Summary

- **RCLL** and **RCLL\*** show practically no violations across any simulation or sample size.
- **SBS** shows time- and sample-size-dependent violations, mostly at smaller $n$, especially for early evaluation times ($\tau^* = q_{0.1}$).
All differences are small (typically < 0.01).
- **ISBS** showed minor violations only at $n \le 50$; none for $n > 100$.
:::

## SBS Improperness: Bias and Tail Effects {-#sbs-bias}

The theoretical shift (i.e. bias) of the global minimum $x^*$ of the expected SBS from the true survival probability at $\tau^*$ is (for $\varepsilon \ge 0$):

$$x^\star - S_Y(\tau^*) = - \frac{\varepsilon (1-S_Y(\tau^*))}{S_C(\tau^*) - \varepsilon} \approx - \varepsilon \cdot \frac{1 - S_Y(\tau^*)}{S_C(\tau^*)}$$

Filter SBS simulation data for plotting:
```{r}
plot_data = data |>
  filter(grepl("^SBS", metric)) |>
  pivot_longer(
    cols = starts_with("shift_"),
    names_to = "shift_type",
    values_to = "shift",
    values_drop_na = TRUE  # only keep the relevant shift per metric
  ) |>
  mutate(
    S_Y = case_when(
      metric == "SBS (Q10)" ~ S_Y_q10,
      metric == "SBS (Median)" ~ S_Y_med,
      metric == "SBS (Q90)" ~ S_Y_q90
    ),
    S_C = case_when(
      metric == "SBS (Q10)" ~ S_C_q10,
      metric == "SBS (Median)" ~ S_C_med,
      metric == "SBS (Q90)" ~ S_C_q90
    )
  ) |>
  select(-S_Y_q10, -S_Y_med, -S_Y_q90, -S_C_q10, -S_C_med, -S_C_q90) |>
  mutate(
    metric = recode(
      metric,
      "SBS (Q10)"   = "SBS (Early, q10)",
      "SBS (Median)"= "SBS (Median, q50)",
      "SBS (Q90)"   = "SBS (Late, q90)"
    )
  )
```

::: {.panel-tabset}

### $\hat \varepsilon$ vs bias
```{r}
#| fig-cap: "Relationship between survival tail product estimate $\\hat{\\varepsilon} = \\frac{\\sum_{d=1}^{1000} \\left[S_Y^d(t_{\\max}^d) \\times S_C^d(t_{\\max}^d)\\right]}{1000}$ and theoretical shift from the global minimum, using a sample of data from the Weibull simulation experiments"
#| cache: true
#| fig-width: 9
#| fig-height: 5

ann_df = data.frame(
  metric = unique(plot_data$metric)[1],
  x = 0.01,
  y = -0.07,
  xend = 0.01,
  yend = -0.175,
  label = "larger bias\n(improperness)"
)

# shift vs epsilon
set.seed(42)
plot_data |>
  slice_sample(prop = 0.2) |> # 20% of: 3 SBS metrics x 10 sample sizes (n) x 10000 simulations per n
  ggplot(aes(x = epsilon, y = shift, color = factor(n))) +
    geom_point(alpha = 0.5, size = 0.1) +
    facet_wrap(~metric, scales = "fixed") +
    scale_color_discrete(
      name = "Sample size (n)",
    ) +
    scale_x_continuous(
      limits = c(0, 0.11),
      breaks = c(0, 0.025, 0.05, 0.075, 0.1)
    ) +
    labs(
      x = "Tail product estimate ε̂ ", #expression(hat(epsilon) * " "),
      y = expression(x^"*" ~ "-" ~ S[Y](tau^ ~ symbol("\052"))), # shift from optimal prediction
      color = "Sample size (n)"
    ) +
    theme_bw(base_size = 14, base_family = "Arial") +
    #theme(axis.text.x = element_text(angle = 35, hjust = 1)) +
    guides(color = guide_legend(override.aes = list(size = 4))) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3, color = "black") +
    geom_segment(
      data = ann_df,
      aes(x = x, y = y, xend = xend, yend = yend),
      inherit.aes = FALSE,linewidth = 1,
      arrow = arrow(length = unit(0.25, "cm"))
    ) +
    geom_text(
      data = ann_df,
      aes(x = x + 0.02, y = y - 0.05, label = label),
      inherit.aes = FALSE,
      angle = 90,
      size = 6,
      fontface = "bold"
    )
#ggsave(filename = "epsilon_vs_bias.png", dpi = 600, height = 5, width = 10, units = "in")
```

### $S_Y(\tau^*)$ vs bias
```{r}
#| fig-cap: "Relationship between true survival $S_Y(\\tau^*)$ and empirical shift from the global minimum, using a sample of data from the Weibull simulation experiments"
#| cache: true
#| fig-width: 9
#| fig-height: 5

set.seed(42)
plot_data |>
  slice_sample(prop = 0.2) |> # 3 SBS metrics x 10 sample sizes (n) x 10000 simulations per n
  ggplot(aes(x = S_Y, y = shift, color = factor(n))) +
  geom_point(alpha = 0.5, size = 0.3) +
  facet_wrap(~metric, scales = "fixed") +
  scale_color_discrete(name = "Sample size (n)") +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = c(0, 0.25, 0.5, 0.75, 1)
  ) +
  labs(
    x = expression(S[Y](tau^ ~ symbol("\052"))),
    y = expression(x^"*" ~ "-" ~ S[Y](tau^ ~ symbol("\052"))),
  ) +
  theme_bw(base_size = 14) +
  theme(panel.spacing = unit(1, "lines")) +
  guides(color = guide_legend(override.aes = list(size = 4))) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3, color = "black") +
  geom_segment(
    data = ann_df,
    aes(x = x, y = y, xend = xend, yend = yend),
    inherit.aes = FALSE,linewidth = 1,
    arrow = arrow(length = unit(0.25, "cm"))
  ) +
  geom_text(
    data = ann_df,
    aes(x = x + 0.17, y = y - 0.05, label = label),
    inherit.aes = FALSE,
    angle = 90,
    size = 6,
    fontface = "bold"
  )
```

### $S_C(\tau^*)$ vs bias
```{r}
#| fig-cap: "Relationship between censoring survival $S_C(\\tau^*)$ and empirical shift from the global minimum, using a sample of data from the Weibull simulation experiments"
#| cache: true
#| fig-width: 9
#| fig-height: 5

set.seed(42)
plot_data |>
  slice_sample(prop = 0.2) |> # 3 SBS metrics x 10 sample sizes (n) x 10000 simulations per n
  ggplot(aes(x = S_C, y = shift, color = factor(n))) +
  geom_point(alpha = 0.5, size = 0.3) +
  facet_wrap(~metric, scales = "fixed") +
  scale_color_discrete(name = "Sample size (n)") +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = c(0, 0.25, 0.5, 0.75, 1)
  ) +
  labs(
    x = expression(S[C](tau^ ~ symbol("\052"))),
    y = expression(x^"*" ~ "-" ~ S[Y](tau^ ~ symbol("\052"))),
  ) +
  theme_bw(base_size = 14) +
  theme(panel.spacing = unit(1, "lines")) +
  guides(color = guide_legend(override.aes = list(size = 4))) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3, color = "black") +
  geom_segment(
    data = ann_df,
    aes(x = x, y = y, xend = xend, yend = yend),
    inherit.aes = FALSE,linewidth = 1,
    arrow = arrow(length = unit(0.25, "cm"))
  ) +
  geom_text(
    data = ann_df,
    aes(x = x + 0.17, y = y - 0.05, label = label),
    inherit.aes = FALSE,
    angle = 90,
    size = 6,
    fontface = "bold"
  )
```

<!-- ### Ratio term vs bias ($S_Y$ color) -->
```{r}
#| fig-cap: "Relationship between $(1-S_Y(\\tau^*))/S_C(\\tau^*)$ and theoretical shift from the global minimum, using a sample of data from the Weibull simulation experiments (n = 100, with $\\varepsilon$ ~ 0.01). Color annotation: $S_Y(\\tau^*)$"
#| cache: true
#| fig-width: 10
#| fig-height: 5
#| eval: false
#| include: false

temp_data = plot_data |>
  filter(n == 100) |> # epsilon ~ 0.01
  mutate(approx_term = (1 - S_Y)/S_C) |>
  # Bin S_Y and S_C
  mutate(
    S_Y_bin = cut(
      S_Y,
      #breaks = quantile(S_C, probs = seq(0, 1, 0.25), na.rm = TRUE),
      breaks = seq(0, 1, by = 0.2),
      include.lowest = TRUE,
      labels = c("[0,0.2)", "[0.2,0.4)", "[0.4,0.6)", "[0.6,0.8)", "[0.8,1]")
    ),
    S_C_bin = cut(
      S_C,
      breaks = seq(0, 1, by = 0.2),
      include.lowest = TRUE,
      labels = c("[0,0.2)", "[0.2,0.4)", "[0.4,0.6)", "[0.6,0.8)", "[0.8,1]")
    )
  )

ann_df = data.frame(
  metric = unique(temp_data$metric)[1],
  x = 0.03,
  y = -0.01,
  xend = 0.03,
  yend = -0.03,
  label = "more improper"
)

temp_data |>
  ggplot(aes(x = approx_term, y = shift, color = S_Y_bin)) +
  geom_point(alpha = 0.4, size = 0.2) +
  facet_wrap(~metric, scales = "fixed") +
  scale_color_manual(
    values = c(
      "#2166AC",  # dark blue
      "#67A9CF",  # light blue
      "#BDBDBD",  # grey
      "#EF8A62",  # light red
      "#B2182B"   # dark red
    ),
    name = expression(S[Y](tau^ ~ symbol("\052")))
  ) +
  labs(
    x = expression((1 - S[Y](tau^ ~ symbol("\052")))/S[C](tau^ ~ symbol("\052"))),
    y = expression(x^"*" ~ "-" ~ S[Y](tau^ ~ symbol("\052")))
  ) +
  theme_bw(base_size = 14) +
  theme(panel.spacing = unit(1, "lines")) +
  guides(color = guide_legend(override.aes = list(size = 8))) +
  geom_segment(
    data = ann_df,
    aes(x = x, y = y, xend = xend, yend = yend),
    inherit.aes = FALSE,linewidth = 1,
    arrow = arrow(length = unit(0.25, "cm"))
  ) +
  geom_text(
    data = ann_df,
    aes(x = x + 0.2, y = y - 0.01, label = label),
    inherit.aes = FALSE,
    angle = 90,
    size = 6,
    fontface = "bold"
  )
```

<!-- ### Ratio term vs bias ($S_C$ color) -->
```{r}
#| fig-cap: "Relationship between $(1-S_Y(\\tau^*))/S_C(\\tau^*)$ and theoretical shift from the global minimum, using a sample of data from the Weibull simulation experiments (n = 100, with $\\varepsilon$ ~ 0.01). Color annotation: $S_C(\\tau^*)$"
#| cache: true
#| fig-width: 10
#| fig-height: 5
#| eval: false
#| include: false

temp_data |>
  ggplot(aes(x = approx_term, y = shift, color = S_C_bin)) +
  geom_point(alpha = 0.4, size = 0.2) +
  facet_wrap(~metric, scales = "fixed") +
  scale_color_manual(
    values = c(
      "#2166AC",  # dark blue
      "#67A9CF",  # light blue
      "#BDBDBD",  # grey
      "#EF8A62",  # light red
      "#B2182B"   # dark red
    ),
    name = expression(S[C](tau^ ~ symbol("\052")))
  ) +
  labs(
    x = expression((1 - S[Y](tau^ ~ symbol("\052")))/S[C](tau^ ~ symbol("\052"))),
    y = expression(x^"*" ~ "-" ~ S[Y](tau^ ~ symbol("\052")))
  ) +
  theme_bw(base_size = 14) +
  theme(panel.spacing = unit(1, "lines")) +
  guides(color = guide_legend(override.aes = list(size = 8))) +
  geom_segment(
    data = ann_df,
    aes(x = x, y = y, xend = xend, yend = yend),
    inherit.aes = FALSE,linewidth = 1,
    arrow = arrow(length = unit(0.25, "cm"))
  ) +
  geom_text(
    data = ann_df,
    aes(x = x + 0.2, y = y - 0.01, label = label),
    inherit.aes = FALSE,
    angle = 90,
    size = 6,
    fontface = "bold"
  )
```

### Bias surface over ($S_Y,S_C$) (n = 50)
```{r}
#| fig-cap: "Interpolated surface of empirical shift $x^\\star - S_Y(\\tau^*)$ as a function of $S_Y(\\tau^*)$ and $S_C(\\tau^*)$ for $n = 50$ (with $\\hat{\\varepsilon}$ ~ 0.02). Panels correspond to early, median, and late evaluation times."
#| cache: true
#| fig-width: 10
#| fig-height: 5

base_data = plot_data |>
  dplyr::filter(n == 50)

interp_shift = function(df, nx = 150, ny = 150) {
  if (nrow(df) == 0) return(NULL)

  interp_res = with(df,
    akima::interp(x = S_C, y = S_Y, z = shift, nx = nx, ny = ny, duplicate = "mean")
  )

  expand.grid(S_C = interp_res$x, S_Y = interp_res$y) |>
    mutate(shift = as.vector(interp_res$z),
           metric = unique(df$metric))
}

interp_list = base_data |>
  dplyr::group_split(metric, .keep = TRUE) |>
  purrr::map(.f = interp_shift)

surface_data = bind_rows(interp_list) |>
  dplyr::filter(!is.na(shift))

range_shift = range(surface_data$shift, na.rm = TRUE)

min_shift = range_shift[1] # most negative
max_shift = range_shift[2] # closest to zero

ggplot(surface_data, aes(x = S_C, y = S_Y, fill = shift)) +
  geom_tile() +
  facet_wrap(~ metric) +
  scale_fill_gradientn(
    colours = c("#B2182B", "grey80", "#2166AC"),
    values = scales::rescale(c(min_shift, (min_shift + max_shift)/2, max_shift)),
    limits = c(min_shift, 0), # Extend upper limit to 0
    name = expression(x^"*" ~ "-" ~ S[Y](tau^ ~ symbol("\052")))
  ) +
  labs(
    x = expression(S[C](tau^ ~ symbol("\052"))),
    y = expression(S[Y](tau^ ~ symbol("\052")))
  ) +
  theme_bw(base_size = 14) +
  theme(panel.spacing = unit(1, "lines"))
#ggsave(filename = "bias_surface_n50.png", dpi = 600, height = 5, width = 10, units = "in")
```

### Bias surface over ($S_Y,S_C$) (n = 100)
```{r}
#| fig-cap: "Interpolated surface of empirical shift $x^\\star - S_Y(\\tau^*)$ as a function of $S_Y(\\tau^*)$ and $S_C(\\tau^*)$ for $n = 100$ (with $\\hat{\\varepsilon}$ ~ 0.01). Panels correspond to early, median, and late evaluation times."
#| cache: true
#| fig-width: 10
#| fig-height: 5

base_data2 = plot_data |>
  dplyr::filter(n == 100)

interp_list2 = base_data2 |>
  dplyr::group_split(metric, .keep = TRUE) |>
  purrr::map(.f = interp_shift)

surface_data2 = bind_rows(interp_list2) |>
  dplyr::filter(!is.na(shift))

range_shift2 = range(surface_data2$shift, na.rm = TRUE)

min_shift = range_shift2[1] # most negative
max_shift = range_shift2[2] # closest to zero

ggplot(surface_data2, aes(x = S_C, y = S_Y, fill = shift)) +
  geom_tile() +
  facet_wrap(~ metric) +
  scale_fill_gradientn(
    colours = c("#B2182B", "grey80", "#2166AC"),
    values = scales::rescale(c(min_shift, (min_shift + max_shift)/2, max_shift)),
    limits = c(min_shift, 0), # Extend upper limit to 0
    name = expression(x^"*" ~ "-" ~ S[Y](tau^ ~ symbol("\052")))
  ) +
  labs(
    x = expression(S[C](tau^ ~ symbol("\052"))),
    y = expression(S[Y](tau^ ~ symbol("\052")))
  ) +
  theme_bw(base_size = 14) +
  theme(panel.spacing = unit(1, "lines"))
#ggsave(filename = "bias_surface_n100.png", dpi = 600, height = 5, width = 10, units = "in")
```
:::

```{r}
plot_data |>
  group_by(n) |>
  summarise(
    mean_hat_epsilon = mean(epsilon)
  ) |>
  DT::datatable(
    rownames = FALSE,
    width = "52%",
    options = list(dom = "t", searching = FALSE, lengthChange = FALSE),
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: center; font-size:150%",
      "Empirical tail product estimate vs sample size"
    )
  ) |>
  formatRound(columns = 2, digits = 4)
```

## SBS Improperness: administrative censoring {-#sbs-admin}

:::{.callout-note}
Here we investigate the impact of administritive censoring on SBS properness.
**Administritive censoring is applied at the 80% percentile of true event times**.
SBS is calculated at the 10th, 50th and 90th percentile of *observed* times (after application of the administrative censoring).

Use [properness_test_admin_cens.R](https://github.com/survival-org/scoring-rules-2024/blob/main/properness_test_admin_cens.R) to run the experiment.

Execute: `Rscript properness_test_admin_cens.R 10000 1000 10000 FALSE 42` to run for $n = 10000$.

Simulation results are available [here](https://github.com/survival-org/scoring-rules-2024/blob/main/results/res_sims10000_distrs1000_n10000_0_admin_cens.rds).
:::

```{r}
res = readRDS("results/res_sims10000_distrs1000_n10000_0_admin_cens.rds") |>
  select(!matches("shape|scale|prop_cens|tv_dist|n_events|max_t|m_true|m_pred"))

measures = c("SBS_q10", "SBS_median", "SBS_q90", "ISBS")
n_distrs = 1000 # `m` value from paper's experiment
tcrit = qt(0.975, df = n_distrs - 1) # 95% t-test
threshold = 1e-4

data = measures |>
  lapply(function(m) {
    mean_diff_col = paste0(m, "_diff")
    mean_sd_col = paste0(m, "_sd")

    res |>
      select(sim, n, !!mean_diff_col, !!mean_sd_col, starts_with("shift_"),
             epsilon, starts_with("S_")) |>
      select(-ends_with("tail")) |>
      mutate(
        se = .data[[mean_sd_col]] / sqrt(n_distrs),
        CI_lower = .data[[mean_diff_col]] - tcrit * se,
        CI_upper = .data[[mean_diff_col]] + tcrit * se,
        signif_violation = .data[[mean_diff_col]] > threshold & CI_lower > 0,
        # Keep relevant SBS columns
        shift_q10 = if (m == "SBS_q10") shift_q10 else NA_real_,
        S_Y_q10   = if (m == "SBS_q10") S_Y_q10 else NA_real_,
        S_C_q10   = if (m == "SBS_q10") S_C_q10 else NA_real_,
        shift_med = if (m == "SBS_median") shift_med else NA_real_,
        S_Y_med   = if (m == "SBS_median") S_Y_med else NA_real_,
        S_C_med   = if (m == "SBS_median") S_C_med else NA_real_,
        shift_q90 = if (m == "SBS_q90") shift_q90 else NA_real_,
        S_Y_q90   = if (m == "SBS_q90") S_Y_q90 else NA_real_,
        S_C_q90   = if (m == "SBS_q90") S_C_q90 else NA_real_
      ) |>
      select(!se) |>
      mutate(metric = !!m) |>
      relocate(metric, .after = 1) |>
      rename(diff = !!mean_diff_col,
             sd   = !!mean_sd_col)
  }) |>
  bind_rows()

data$metric = factor(
  data$metric,
  levels = measures,
  labels = c("SBS (Early, q10)", "SBS (Median, q50)", "SBS (Late, q90)", "ISBS")
)

admin_all_stats =
  data |>
  group_by(n, metric) |>
  summarize(
    n_violations = sum(signif_violation),
    violation_rate = mean(signif_violation),
    mean_hat_epsilon = mean(epsilon),
    diff_mean = if (any(signif_violation)) mean(diff[signif_violation]) else NA_real_,
    diff_median = if (any(signif_violation)) median(diff[signif_violation]) else NA_real_,
    diff_min = if (any(signif_violation)) min(diff[signif_violation]) else NA_real_,
    diff_max = if (any(signif_violation)) max(diff[signif_violation]) else NA_real_,
    .groups = "drop"
  )

admin_all_stats |>
  arrange(metric) |>
  DT::datatable(
    rownames = FALSE,
    options = list(dom = "t", searching = FALSE),
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: center; font-size:142%",
      "SBS: Empirical violations of properness for large sample size + administrative censoring")
  ) |>
  formatRound(columns = 4:9, digits = 5)
```

## Table C1 {-#tabc1}

:::{.callout-note}
The only difference with the [previous experiment](#tab1) is that $S_C(t)$ is now being estimated using the marginal Kaplan-Meier model via `survival::survfit()` instead of using the true Weibull censoring distribution (see [helper.R](https://github.com/survival-org/scoring-rules-2024/blob/main/helper.R)).
For SBS/ISBS we use *constant interpolation* of the censoring survival distribution $S_C(t)$.
For RCLL\* we use *linear interpolation* of $S_C(t)$ to mitigate density estimation issues, i.e. for $f_C(t)$.

Use [properness_test.R](https://github.com/survival-org/scoring-rules-2024/blob/main/properness_test.R) to run the experiments.
To run for different sample sizes `n`, use [run_tests.sh](https://github.com/survival-org/scoring-rules-2024/blob/main/run_tests.sh), changing `estimate_cens` to `TRUE`.
Merged simulation results for all `n` are available [here](https://github.com/survival-org/scoring-rules-2024/blob/main/results/res_sims10000_distrs1000_1.rds).
:::

Load results (same output columns as in the previous section):
```{r}
res = readRDS("results/res_sims10000_distrs1000_1.rds") |>
  select(!matches("shape|scale|prop_cens|tv_dist|n_events|max_t|m_true|m_pred")) # remove columns
```

As before, we compute 95% confidence intervals for the score differences across $m = 1000$ draws per simulation:
```{r}
measures = c("SBS_q10", "SBS_median", "SBS_q90", "ISBS", "RCLL", "wRCLL")
n_distrs = 1000 # `m` value from paper's experiment
tcrit = qt(0.975, df = n_distrs - 1) # 95% t-test
threshold = 1e-4

data = measures |>
  lapply(function(m) {
    mean_diff_col = paste0(m, "_diff")
    mean_sd_col = paste0(m, "_sd")

    res |>
      select(sim, n, !!mean_diff_col, !!mean_sd_col, starts_with("shift_"),
             epsilon, starts_with("S_")) |>
      select(-ends_with("tail")) |>
      mutate(
        se = .data[[mean_sd_col]] / sqrt(n_distrs),
        CI_lower = .data[[mean_diff_col]] - tcrit * se,
        CI_upper = .data[[mean_diff_col]] + tcrit * se,
        signif_violation = .data[[mean_diff_col]] > threshold & CI_lower > 0,
        # Keep relevant SBS columns, just in case
        shift_q10 = if (m == "SBS_q10") shift_q10 else NA_real_,
        S_Y_q10   = if (m == "SBS_q10") S_Y_q10 else NA_real_,
        S_C_q10   = if (m == "SBS_q10") S_C_q10 else NA_real_,
        shift_med = if (m == "SBS_median") shift_med else NA_real_,
        S_Y_med   = if (m == "SBS_median") S_Y_med else NA_real_,
        S_C_med   = if (m == "SBS_median") S_C_med else NA_real_,
        shift_q90 = if (m == "SBS_q90") shift_q90 else NA_real_,
        S_Y_q90   = if (m == "SBS_q90") S_Y_q90 else NA_real_,
        S_C_q90   = if (m == "SBS_q90") S_C_q90 else NA_real_
      ) |>
      select(!se) |>
      mutate(metric = !!m) |>
      relocate(metric, .after = 1) |>
      rename(diff = !!mean_diff_col,
             sd   = !!mean_sd_col)
  }) |>
  bind_rows()

data$metric = factor(
  data$metric,
  levels = measures,
  labels = c("SBS (Q10)", "SBS (Median)", "SBS (Q90)", "ISBS", "RCLL", "RCLL*")
)
```

Lastly, we summarize the significant violations across sample sizes and scoring rules:
```{r}
all_stats =
  data |>
  group_by(n, metric) |>
  summarize(
    n_violations = sum(signif_violation),
    violation_rate = mean(signif_violation),
    diff_mean = if (any(signif_violation)) mean(diff[signif_violation]) else NA_real_,
    diff_median = if (any(signif_violation)) median(diff[signif_violation]) else NA_real_,
    diff_min = if (any(signif_violation)) min(diff[signif_violation]) else NA_real_,
    diff_max = if (any(signif_violation)) max(diff[signif_violation]) else NA_real_,
    .groups = "drop"
  )

all_stats |>
  arrange(metric) |>
  DT::datatable(
    rownames = FALSE,
    options = list(searching = FALSE),
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: center; font-size:150%",
      "Table C1: Empirical violations of properness using G(t)")
  ) |>
  DT::formatRound(columns = 4:8, digits = 5)
```

:::{.callout-tip}
## Summary

Estimating $S_C(t)$ via Kaplan-Meier produced nearly identical results to using the true censoring distribution, with no violations for RCLL/RCLL* and only small-sample effects for SBS and ISBS.
:::

## Sensitivity to Model Misspecification {-#sen}

To evaluate how effectively each scoring rule discriminates between increasingly misspecified models, we designed a simulation study based on a non-proportional hazards data-generating process (DGP).
The DGP is a log-normal accelerated failure time (AFT) model with covariate-dependent location and group-specific scale, inducing crossing survival curves.
For more information about the DGP, generated tasks, models and benchmark details, please see the manuscript.

:::{.callout-note}
The simulation and benchmarking pipeline is implemented in the [simulation_benchmark](https://github.com/survival-org/scoring-rules-2024/tree/main/simulation_benchmark) directory.
Key scripts include:

- [simulate.R](https://github.com/survival-org/scoring-rules-2024/blob/main/simulation_benchmark/simulate.R): Defines the DGP and can generate data sets under **low** and **high** censoring settings.
- [train.R](https://github.com/survival-org/scoring-rules-2024/blob/main/simulation_benchmark/train.R): Trains all models on the generated tasks. The trained models are saved as `trained_low.rds`, `trained_high.rds`, and separate files for the large Random Survival Forest objects (`trained_low_rsf.rds`, `trained_high_rsf.rds`).
- [execute_dense.R](https://github.com/survival-org/scoring-rules-2024/blob/main/simulation_benchmark/execute_dense.R): Runs a benchmark where the specified models are evaluated on test sets of varying sizes using a dense prediction time grid (unique event times from the training task). Results are saved in [`results/sim_bm_dense.rds`](https://github.com/survival-org/scoring-rules-2024/blob/main/results/sim_bm_dense.rds).
- [RCLL_investigation.R](https://github.com/survival-org/scoring-rules-2024/blob/main/simulation_benchmark/RCLL_investigation.R): Investigates the sensitivity of the RCLL to the resolution of the prediction time grid. Results are saved in [`results/rcll_sim.rds`](https://github.com/survival-org/scoring-rules-2024/blob/main/results/rcll_sim.rds).
:::

### DGP and censoring scenarios

The following code produces Kaplan–Meier curves for both tasks with additional summary statistics, stratified by the binary treatment variable.
See Figure 4 in the paper.

```{r}
#| cache: true
#| warning: false
tasks = readRDS("simulation_benchmark/tasks.rds")

plot_km_task = function(task, title, show_censor = FALSE, extend_to_last_time = TRUE) {
  d = task$data()

  # summary quantities
  cens_total = mean(d$status == 0)
  cens_x20 = mean(d$status == 0 & d$x2 == 0)
  cens_x21 = mean(d$status == 0 & d$x2 == 1)
  n_events = sum(d$status == 1)
  admin_cens = mean(d$time == 10)

  subtitle =
    paste0(
      sprintf("Total censoring: %.1f%% (x2 = 0: %.1f%%, x2 = 1: %.1f%%)\n",
        100 * cens_total, 100 * cens_x20, 100 * cens_x21
      ),
      sprintf("Number of events: %d\n", n_events),
      sprintf("Administrative censoring: %.1f%%", 100*admin_cens)
    )

  # Kaplan-Meier fit
  fit = survival::survfit(
    survival::Surv(time, status) ~ factor(x2),
    data = d
  )

  s = summary(fit)
  df = data.frame(
    time = s$time,
    surv = s$surv,
    lower = s$lower,
    upper = s$upper,
    strata = s$strata,
    n_censor = s$n.censor
  )

  # ensure S(0) = 1 for each strata
  by_strata = split(df, df$strata)
  s0_rows = lapply(by_strata, function(dd) {
    if (any(dd$time == 0)) {
      NULL
    } else {
      data.frame(
        time = 0,
        surv = 1,
        lower = 1,
        upper = 1,
        strata = dd$strata[1],
        n_censor = 0
      )
    }
  })
  s0_rows = do.call(rbind, s0_rows)
  if (!is.null(s0_rows)) {
    df = rbind(s0_rows, df)
  }

  if (extend_to_last_time) {
    max_time = max(d$time)
    by_strata = split(df, df$strata)
    ext = lapply(by_strata, function(dd) {
      last_row = dd[nrow(dd), , drop = FALSE]
      if (max_time > last_row$time) {
        last_row$time = max_time
        last_row
      } else {
        NULL
      }
    })
    ext = do.call(rbind, ext)
    if (!is.null(ext)) {
      df = rbind(df, ext)
    }
  }

  censor_df = df[df$n_censor > 0, , drop = FALSE]

  p =
    ggplot(df, aes(x = time, y = surv, color = strata, linetype = strata)) +
    geom_ribbon(
      aes(ymin = lower, ymax = upper, group = strata),
      fill = "grey80",
      alpha = 0.4,
      color = NA
    ) +
    geom_step() +
    scale_color_manual(
      values = c("black", "red"),
      labels = c(
        expression(x[2] == 0 ~ paste("(", sigma, "  = ", 0.5, ")")),
        expression(x[2] == 1 ~ paste("(", sigma, "  = ", 1.5, ")"))
      )
    ) +
    scale_linetype_manual(
      values = c(1, 1),
      labels = c(
        expression(x[2] == 0 ~ paste("(", sigma, "  = ", 0.5, ")")),
        expression(x[2] == 1 ~ paste("(", sigma, "  = ", 1.5, ")"))
      )
    ) +
    labs(
      title = title,
      subtitle = subtitle,
      x = "Time",
      y = "Survival probability",
      color = NULL,
      linetype = NULL
    ) +
    theme_bw(base_size = 14) +
    theme(
      legend.position = c(0.98, 0.98),
      legend.justification = c(1, 1),
      legend.background = element_rect(fill = "white", color = "grey80"),
      legend.text = element_text(size = 16),
      legend.title = element_text(size = 16),
      legend.key.size = grid::unit(1, "lines")
    )

  if (show_censor && nrow(censor_df) > 0) {
    p = p + geom_point(
      data = censor_df,
      aes(x = time, y = surv, color = strata),
      shape = 3,
      stroke = 0.9,
      size = 2,
      show.legend = FALSE
    )
  }

  p
}
```

:::{.panel-tabset}

### Low censoring DGP
```{r}
#| warning: false
#| fig-width: 8
#| fig-height: 6
p_low  = plot_km_task(tasks$low, title = "Low censoring task", show_censor = TRUE)
p_low
```

### High censoring DGP
```{r}
#| warning: false
#| fig-width: 8
#| fig-height: 6
p_high = plot_km_task(tasks$high, title = "High censoring task", show_censor = TRUE)
p_high
```

```{r}
#| include: false
#| eval: false
(p_low | p_high) # combine using patchwork
ggsave("kaplan_meier_plots_DPG.png", device = png, dpi = 600, width = 11, height = 6)
```
:::

### Large sample size sensitivity

We now evaluate how effectively each scoring rule discriminates between correctly specified and increasingly misspecified models under a large test set.
For each task, we generate 100 Monte Carlo resamplings (test sets) of size $n = 1000$.
For each replicate, we compute the excess loss, i.e. the difference in empirical loss between a candidate model and the true data-generating process, i.e. $\Delta L = L(\hat S) - L(S_Y)$, and plot it against the mean integrated absolute error (MIAE) between the predicted and true survival functions (which is a measure of misspecification).

The panels below panels corresponding to three scoring rules (left to right):

- **RCLL**: right-censored log-loss (density estimated via linear interpolation of the survival function)
- **SBS**: Survival Brier Score evaluated at $t = 5$
- **ISBS**: Integrated Survival Brier Score, integrated up to the time corresponding to the $80\%$ proportion of censoring in each test set
- **Harrell's C-index**: a standard discrimination metric (i.e. *not* a scoring rule) which is not proper as we can see below for the high censoring task

```{r}
#| cache: true
res = readRDS("results/sim_bm_dense.rds")

# Plot score differences vs. MIAE/MISE for multiple measures and one test size value
# Score diffs are (predicted - true)
plot_score_diff = function(
  res,
  y = "miae", # "mise" or "miae"
  task_id_val = "low", # "low" or "high"
  n_test_val = 1000, # test set size to display
  drop_learners = c("CoxPH"), # remove undesired learners
  rcll_truth = "interp", # "true" => (true S, true f), "interp" => (true S, interpolated f from linear S)
  meas = c("rcll", "cindex", "sbs", "isbs"), # measures to display
  legend_position = "right" # "right", "left", "top", "bottom", "none"
) {
  # Subset to the chosen task
  dt = res[task_id == task_id_val & n_test == n_test_val]
  # Remove unwanted learners
  dt = dt[!learner_id %in% drop_learners]

  # Rename learner_ids
  label_map = c(
    "LogNorm_int_shape_x2"   = "Oracle",
    "LogLog_int_shape_x2"    = "LLogis",
    "Weib_int_shape_x2"      = "Weibull",
    "LogNorm_noint_shape_x2" = "LogNorm_scale",
    "LogNorm_int_noshape"    = "LogNorm_int",
    "LogNorm_noint_noshape"  = "LogNorm",
    "CoxPH_int"              = "Cox_int",
    "RSF"                    = "RSF",
    "KM"                     = "KM"
  )

  dt[, learner_id := ifelse(
    learner_id %in% names(label_map),
    label_map[learner_id],
    learner_id
  )]

  # --- Extract true model references per replicate ---
  true_dt = dt[learner_id == "true",
    .(rsmp_id,
      rcll_true_full = true_rcll,
      rcll_true_interp = rcll,
      true_cindex = cindex,
      true_sbs = sbs,
      true_isbs = isbs)]

  dt = merge(dt, true_dt, by = "rsmp_id")

  # --- Compute differences ---
  # RCLL: compare to either full true Likelihood or f interpolated from linear S
  if (rcll_truth == "interp") {
    dt[, diff_rcll := rcll - rcll_true_interp]
  } else if (rcll_truth == "true") {
    # comparing with the true model's RCLL, makes the rest of the
    # metric comparisons a bit less comparable (as we don't have the true ISBS or SBS)
    dt[, diff_rcll := rcll - rcll_true_full]
  }

  # C‑index: error = 1 - cindex
  # error_pred - error_true = (1 - cindex_pred) - (1 - cindex_true) = cindex_true - cindex_pred
  dt[, diff_cindex := true_cindex - cindex]

  # Brier score
  dt[, diff_sbs := sbs - true_sbs]

  # Integrated Brier score
  dt[, diff_isbs := isbs - true_isbs]

  # Remove true model
  dt = dt[learner_id != "true"]

  # --- Reshape to long format ---
  measures = c("rcll", "cindex", "sbs", "isbs")
  measures = intersect(measures, meas)

  long_dt = melt(dt,
                 id.vars = c("learner_id", "mise", "miae", "rsmp_id"),
                 measure.vars = paste0("diff_", measures),
                 variable.name = "measure",
                 value.name = "diff")
  long_dt[, measure := sub("diff_", "", measure)]

  # Learner colours
  cols = c(
    Oracle        = "#dfce0cff",
    LogNorm_scale = "#2171B5",
    LogNorm_int   = "#6BAED6",
    LogNorm       = "#C6DBEF",
    LLogis        = "#FFA500",
    Weibull       = "#756BB1",
    Cox_int       = "#238B45",
    RSF           = "#D62728",
    KM            = "#7F7F7F"
  )

  # Keep only colours that appear in the data
  present_learners = unique(long_dt$learner_id)
  cols = cols[names(cols) %in% present_learners]

  # Label measures nicely
  measure_labels = c(
    rcll   = if (rcll_truth == "interp") "RCLL" else "RCLL (difference to true likelihood)",
    sbs    = "SBS (t = 5)",
    isbs   = "ISBS (τ* = q90)",
    cindex = "1 − C‑index (error)"
  )
  measure_labels = measure_labels[names(measure_labels) %in% measures]
  long_dt[, measure_label := factor(measure,
                                    levels = names(measure_labels),
                                    labels = measure_labels)]

  # Order learners for the legend
  learner_order = c(
    "Oracle",
    "LLogis",
    "Weibull",
    "LogNorm_scale",
    "LogNorm_int",
    "LogNorm",
    "Cox_int",
    "RSF",
    "KM"
  )
  learner_order = learner_order[learner_order %in% present_learners]
  long_dt[, learner_id := factor(learner_id, levels = learner_order)]

  # Create the plot
  ggplot(long_dt, aes(x = .data[[y]], y = diff, color = learner_id)) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5) +
    geom_point(alpha = 0.6, size = 1.5) +
    scale_color_manual(values = cols) +
    facet_wrap(~ measure_label, scales = "free_y") +
    labs(
      title = sprintf(
        "%s censoring task (Test set size = %d)", # unique event times as grid by default
        paste0(toupper(substring(task_id_val, 1, 1)), substring(task_id_val, 2)),
        n_test_val
      ),
      x = ifelse(y == "mise",
                 "MISE (distance to true S)",
                 "MIAE (distance to true S)"),
      #y = "Score difference (predicted − true model)",
      y = expression(L[pred] - ~L[true]),
      color = "Model"
    ) +
    theme_bw(base_size = 12, base_family = "Arial") +
    theme(
      strip.background = element_rect(fill = "grey95"),
      strip.text = element_text(face = "bold"),
      legend.position = if (legend_position == "none") "none" else legend_position,
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 12)
    ) +
    guides(color = guide_legend(override.aes = list(size = 5)))
}
```

:::{.panel-tabset}

### Low censoring DGP
```{r}
#| fig-width: 9
#| fig-height: 6
plot_score_diff(res, task_id_val = "low", meas = c("rcll", "sbs", "isbs", "cindex"),
                n_test_val = 1000, legend_position = "bottom")
```

### High censoring DGP
```{r}
#| fig-width: 9
#| fig-height: 6
plot_score_diff(res, task_id_val = "high", meas = c("rcll", "sbs", "isbs", "cindex"),
                n_test_val = 1000, legend_position = "bottom")
```

```{r}
#| include: false
#| eval: false
# figure 5
p1 = plot_score_diff(res, task_id_val = "low", meas = c("rcll", "sbs", "isbs"),
                     n_test_val = 1000, legend_position = "none")
p2 = plot_score_diff(res, task_id_val = "high", meas = c("rcll", "sbs", "isbs"),
                     n_test_val = 1000, legend_position = "bottom") +
     guides(color = guide_legend(override.aes = list(size = 5, alpha = 1), nrow = 1, byrow = TRUE)) +
     theme(legend.direction = "horizontal")
(p1 / p2)
ggsave(filename = "sensitivity_large_n.png", dpi = 600, height = 8, width = 11, units = "in")
ggsave(filename = "sensitivity_large_n.pdf", dpi = 600, height = 8, width = 11, units = "in", device = cairo_pdf)
```
:::

:::{.callout-tip}
## Summary

- **RCLL** provides the best discrimination: point clouds are well separated by model, closely track the MIAE ordering, and show low variability. All misspecified models yield positive excess loss confirming empirical properness.
- **SBS** (at $t = 5$) shows poor discrimination: substantial overlap between models, weak alignment with MIAE, and high variability, particularly under high censoring. It is unsuitable for model comparison.
- **ISBS** preserves the global MIAE ranking but tends to form a small number of coarse clusters of models with similar excess loss, within which models are largely indistinguishable. Occasional negative $\Delta L$ values indicate finite‑sample properness violations, especially under high censoring.
:::

### Effect of Test Set Size

Performance of RCLL and ISBS as test set size decreases.
Assesses how sample size affects discriminatory power and empirical properness.
Relates to Figure 6 (low censoring task) and Figure C1 (high censoring task) in the paper.

### Impact of Prediction Grid Resolution on RCLL

Sensitivity of the RCLL to the density of the prediction time grid (subsampling from 2% to 100% of unique event times).
Shows how coarse grids affect bias, variability, and model ranking (related to Figure 7 and Table C2 in the paper).

## Benchmark (Real-World Data) {-#bench}

:::{.callout-note}
This benchmark complements the simulation study by comparing how different evaluation metrics assess and rank survival learners across eight heterogeneous, low-dimensional, real-world datasets.

To reproduce this benchmark, see [real_data_benchmark.R](https://github.com/survival-org/scoring-rules-2024/blob/main/real_data_benchmark.R).
:::

### Datasets

All the datasets are available from `mlr3proba` as `mlr3` tasks.

```{r}
#| message: false
#| warning: false
#| cache: true
library(mlr3proba)
library(mlr3misc)
# silence mlr3 logging messages
lgr::get_logger("mlr3")$set_threshold(0)

res = readRDS("results/real_data_bm.rds")

task_ids = unique(res$task_id)
tasks = mlr3::tsks(task_ids)

metrics = c("C-index", "D-calib", "RCLL*", "RCLL", "ISBS")
res_long = melt(
  res,
  id.vars = c("task_id", "learner_id", "iteration"),
  measure.vars = metrics,
  variable.name = "metric",
  value.name = "value"
)

# remove RCLL* from the boxplots
res_long = res_long[metric != "RCLL*"]

# truncate D-calibration to 50 to avoid extreme outliers in the boxplots
res_long[
  metric == "D-calib",
  value := pmin(value, 50)
]

# Reorder metric factor to the desired sequence
res_long[, metric := factor(metric, levels = metrics)]

task_info = lapply(tasks, function(task) {
  id = task$id
  n = task$nrow
  p = task$n_features
  feature_dt = task$col_info[id %nin% c("..row_id", task$target_names)]
  n_factor = sum(feature_dt$type == "factor")
  n_numeric = sum(feature_dt$type %in% c("numeric", "integer"))
  cens_rate = task$cens_prop()
  admin_cens_rate = task$admin_cens_prop()
  prop_haz = ifelse(task$prop_haz() < 0.05, FALSE, TRUE)

  list(
    id = id,
    n = n,
    p = p,
    n_factor = n_factor,
    n_numeric = n_numeric,
    censoring_rate = cens_rate,
    admin_censoring_rate = admin_cens_rate,
    prop_haz = prop_haz
  )
}) |> rbindlist()
```

In the following table:

- **id** is the dataset identifier
- **n** is the number of observations
- **p** is the number of features
- **n_factor** is the number of factor features
- **n_numeric** is the number of numeric features
- **censoring_rate** is the proportion of censored observations in the dataset
- **admin_censoring_rate** is the proportion of observations censored due to administrative censoring
- **prop_haz** is a boolean derived from the $p$-value of the global proportional hazards test computed with `survival::cox.zph()` at the 5% significance level: `TRUE` indicates no evidence against proportional hazards ($p \ge 0.05$), whereas `FALSE` indicates a violation ($p < 0.05$).

```{r}
task_info |>
  arrange(censoring_rate) |>
  DT::datatable(
    rownames = FALSE,
    filter = "top",
    options = list(dom = "t", searching = TRUE)
  ) |>
  DT::formatRound(columns = c(6, 7), digits = 2) |>
  DT::formatStyle(
    columns = "censoring_rate",
    backgroundColor = DT::styleInterval(
      c(0.25, 0.5, 0.75),
      c("#e5f5e0", "#a1d99b", "#e7785f", "#f8555b")
    )
  )
```

### Models

We used the following models:

- Kaplan-Meier (KM; non-parametric)
- Cox proportional hazards model (Cox; semi-parametric)
- Random survival forests (RSF; non-parametric)
- Log-normal accelerated failure time model (LogNorm; parametric)
- Weibull accelerated failure time model (Weibull; parametric)

Model performance is evaluated using three survival scoring rules (RCLL, RCLL*, and ISBS), Harrell's C-index (discrimination), and D-calibration (calibration).
All models are assessed using 10 repetitions of 5-fold cross-validation without hyperparameter tuning.

### Performance per-dataset

Predictive performance can vary considerably across datasets due to differences in sample size, censoring proportion, covariate distributions and the suitability of the underlying model assumptions.
To provide a detailed overview, the following figures show the distribution of each evaluation metric across all repeated cross-validation resamples, separately for each dataset.
D-calibration is truncated at 50 to avoid extreme outliers in the boxplots.
Within each dataset, learners are ordered by their median RCLL to facilitate comparison across the different evaluation metrics.

```{r}
learner_ids = sort(unique(res_long$learner_id))
cols = scales::hue_pal()(length(learner_ids))
names(cols) = learner_ids
```

```{r}
#| cache: true
#| results: asis
#| fig-width: 10
#| fig-height: 7
plot_task = function(task) {
  dat = res_long[task_id == task]

  # Order learners by median RCLL
  rcll_medians = dat[metric == "RCLL", .(med = median(value)), by = learner_id]
  learner_order = rcll_medians[order(med), learner_id]
  dat[, learner_id := factor(learner_id, levels = learner_order)]

  p = dat |>
    ggplot(aes(
      x = value,
      y = learner_id,
      colour = learner_id,
      fill = learner_id
    )) +
    geom_boxplot(alpha = 0.25, outlier.shape = NA) +
    geom_jitter(width = 0, height = 0.15, alpha = 0.25, size = 0.7) +
    facet_wrap(~metric, scales = "free_x", ncol = 2) +
    # ---- Add the manual scales ----
    scale_colour_manual(values = cols) +
    scale_fill_manual(values = cols) +
    labs(
      x = NULL,
      y = NULL,
      title = task
    ) +
    theme_bw(base_size = 14, base_family = "Arial") +
    theme(
      legend.position = "none",
      # strip.background = element_blank(),
      strip.background = element_rect(fill = "grey95"),
      strip.text = element_text(face = "bold"),
      panel.grid.major.y = element_blank(),
      plot.title = element_text(face = "bold")
    )

  p
}

task_ids = unique(task_info$id)

cat("::: {.panel-tabset}\n\n")

for (task_id in task_ids) {
  title = with(
    task_info[task_info$id == task_id, ],
    sprintf("%s (n=%d, p=%d, %s)", id, n, p, ifelse(prop_haz, "PH", "PH violated"))
  )

  cat("### ", title, "\n\n", sep = "")
  print(plot_task(task_id))
  cat("\n\n")
}

cat(":::\n")
```

For easier numerical comparison, the table below reports the **mean performance of every learner on each dataset**, averaged over the repeated cross-validation resamples.

```{r}
res_mean = res[,
  lapply(.SD, mean, na.rm = TRUE), by = .(task_id, learner_id), .SDcols = metrics
]

res_mean |>
  DT::datatable(
    rownames = FALSE,
    filter = "top",
    options = list(searching = TRUE, pageLength = 10)
  ) |>
  DT::formatRound(columns = c("RCLL", "RCLL*", "C-index", "ISBS"), digits = 3) |>
  DT::formatSignif(columns = "D-calib", digits = 3)
```

### Rank-correlation between evaluation metrics

Absolute performance values may not be directly comparable across datasets, so we evaluate how the different metrics rank competing models within each dataset.
For every dataset and resampling iteration, model performances are converted into ranks (1 = best).
Harrell's C-index is converted to an error metric (1 - C-index) so that all metrics are oriented in the same direction (smaller values indicate better performance).

We compute correlations between the resulting ranking vectors and visualize the empirical agreement between evaluation metrics as a correlation heatmap.

```{r}
#| warning: false
#| message: false
library(ggcorrplot)

# convert C-index to 1 - C
res = res[, `C-index-error` := 1 - `C-index`]

metrics = c("C-index-error", "D-calib", "RCLL*", "RCLL", "ISBS")

# Compute ranks per task for each metric
rank_cols = paste0("rank_", metrics)
res_ranks = copy(res)

# calculate ranks for each metric, grouped by task_id and iteration
for (i in seq_along(metrics)) {
  m = metrics[i]
  res_ranks[,
    paste0("rank_", m) := frank(get(m), ties.method = "min"),
    by = .(task_id, iteration)
  ]
}

cor_all = cor(res_ranks[, ..rank_cols], method = "pearson")

ggcorrplot(
  cor_all,
  method = "square",
  type = "lower",
  hc.order = TRUE,
  lab = TRUE,
  lab_size = 4,
  outline.color = "white"
)
```

:::{.callout-tip}
## Summary

The real-world benchmark confirms that **the choice of evaluation metric can substantially influence the ranking of survival learners.**

Across the eight datasets, RCLL consistently provides the greatest separation between competing models, whereas Harrell's C-index often assigns similar scores to learners with noticeably different predictive distributions.
ISBS exhibits moderate agreement with both RCLL ($\rho=0.43$) and the C-index ($\rho=0.64$).
Across most datasets, ISBS distinguishes models generally less than RCLL.

while D-calibration is largely uncorrelated with the remaining metrics, reflecting its focus on calibration rather than discrimination or overall predictive accuracy.

Interestingly, RCLL and its variant RCLL* exhibit only weak empirical agreement ($\rho=0.16$), illustrating that seemingly similar scoring rules can induce substantially different model rankings in practice.

Taken together, these results complement the simulation study by demonstrating that the empirical behaviour of survival evaluation metrics can differ substantially on real data, reinforcing the importance of selecting evaluation criteria that align with the intended prediction objective.
:::

## R session info {-}

```{r}
utils::sessionInfo()
```
